```{r echo=FALSE}
opts_chunk$set(
    echo = FALSE,
    comment = NA,
    quiet = TRUE,
    progress = FALSE,
    tidy = FALSE,
    cache = FALSE,
    message = FALSE,
    error = FALSE, # FALSE: do not preserve errors. Always stop execution.
    warning = TRUE
)
options(width = 110)
```



# `r settings[["title"]]`


## General settings

_Benthic Ecosystem Quality Index 2 -- package **`r packageDescription("BEQI2", fields = "Package")`** version `r packageDescription("BEQI2", fields = "Version")` (`r packageDescription("BEQI2", fields = "Date")`)_

- Time stamp: `r format(Sys.time())` 
- User name: `r settings[["user"]]`
- Input file: `r settings[["BEQI-file"]]`
- Output file: `r settings[["output-file"]]`
- Log file: `r settings[["log-file"]]`


## Selection of benthos records

```{r}
d_twn <- readTWN(filename = settings[["TWN-file"]])
```

- TWN file: `r settings[["TWN-file"]]`
    - Modification time: `r format(file.info(settings[["TWN-file"]])$mtime)`
    - Number of records read: `r nrow(d_twn)`
    
    
```{r}
d_beqi <- readBEQI(filename = settings[["BEQI-file"]])
```

- Input file: `r settings[["BEQI-file"]]`
    - Number of records read: `r nrow(d_beqi)`

```{r}
# keep only records within the period of interest
month <- as.integer(format(d_beqi$DATE, format = "%m"))
sel <- (month >= settings[["months"]][[1]]) & 
       (month <= settings[["months"]][[2]])
d_beqi <- d_beqi[sel, ]
```

    - Number of records between month `r settings[["months"]][[1]]` and month `r settings[["months"]][[2]]` : `r nrow(d_beqi)`

```{r}
# convert taxa in 'd_beqi' to preferred names
d_beqi$TAXON_NEW <- rename(x = d_beqi$TAXON, from = d_twn$taxonname, to = d_twn$taxon)

# handle azoic (or azoisch) samples (no species available)
index <- grep(pattern = "^Azoi", x = d_beqi$TAXON, ignore.case = TRUE)
d_beqi$TAXON_NEW[index] <- d_beqi$TAXON[index]

# add taxon level and taxon group
index <- match(x = d_beqi$TAXON_NEW, table = d_twn$taxon)
d_beqi$taxonlevel <- d_twn$taxonlevel[index]
d_beqi$taxongroup <- d_twn$taxongroup[index]
```


```{r}
# BEQI2 has been designed for endofauna, therefore remove 
# epifaunal species and insecta

# create black-list of taxon groups to be removed
blacklist <- as.data.frame(
    matrix(data = c(
        "ARACH",    "Arachnida",
        "CRDEC",    "Crustacea - Decapoda",
        "CRMYS",    "Crustacea - Mysida",
        "IDCHI",    "Insecta/Diptera - Chironomidae",
        "IDREM",	"Insecta/Diptera - remaining",
        "IDSIM",	"Insecta/Diptera - Simuliidae",
        "INCOL",	"Insecta - Coleoptera",
        "INEPH",	"Insecta - Ephemeroptera",
        "INHET",	"Insecta - Heteroptera",
        "INLEP",	"Insecta - Lepidoptera",
        "INODO",	"Insecta - Odonata",
        "INREM",	"Insecta - remaining",
        "INTRI",	"Insecta - Trichoptera"
    ), ncol = 2, byrow = TRUE, 
    dimnames = list(NULL, c("taxongroup", "description"))),
    stringsAsFactors = FALSE
)

# merge black-list with BEQI-2 data
d_beqi <- merge(x = d_beqi, y = blacklist, by = "taxongroup", 
                all.x= TRUE, all.y = FALSE, sort = FALSE)

# select all black-list species
blacklisted <- !is.na(d_beqi$description)

if (any(blacklisted)) {
    d <- ddply(
        .data = d_beqi[blacklisted, ], 
        .variables= "description", 
        .fun = function(x) {
            data.frame(frequency = nrow(x))
        }
    )
    names(d)[1] <- "taxon group"
    
    # remove black-list species from BEQI-2 data
    d_beqi <- d_beqi[!blacklisted, ]
}
```

- Total number of non-endofauna records that have been removed: `r if(any(blacklisted)) {sum(d$frequency)} else {0}`
```{r eval=any(blacklisted), results='asis'}
print(xtable(d), type = "html", include.rownames = FALSE)
```


```{r}
# remove CHAR code 'REST'
n <- nrow(d_beqi)
d_beqi <- subset(x = d_beqi, subset = is.na(CHAR) | (toupper(CHAR) != "REST"))
```
- Number of records with CHAR = '`REST`' that have been removed: `r n - nrow(d_beqi)` 


```{r}
# add year
d_beqi$YEAR <- as.integer(format(d_beqi$DATE, format = "%Y"))
```

- Number of records remaining for analysis: `r nrow(d_beqi)` 
- Number of samples remaining for analysis: `r length(unique(d_beqi$ID))`





## Conversion of species names

```{r}
inconvertible <- which(is.na(d_beqi$TAXON_NEW))
```

The following `r length(inconvertible)` taxon names in the BEQI file are inconvertible. These names are not in the Taxa Water management the Netherlands (TWN) list and will be removed:
```{r eval=length(inconvertible)>0L, results='asis'}
d <- ddply(
    .data = d_beqi[inconvertible, ],
    .variables = "TAXON", 
    .fun = function(x) {
        data.frame(
            frequency = nrow(x), 
            similar_names_TWN = toString(
                agrep(
                    pattern = x$TAXON[1], 
                    x = d_twn$taxon, 
                    value = TRUE, 
                    ignore.case = TRUE
                )
            )
        )
    }
)
print(xtable(d), type = "html", include.rownames = FALSE)
d_beqi <- d_beqi[-inconvertible, ]
```




```{r}
d_beqi$TAXON <- d_beqi$TAXON_NEW
d_beqi$TAXON_NEW <- NULL
```

The first column gives the taxon name as found in the BEQI input file (`r basename(settings[["BEQI-file"]])`), the second column gives the number of occurrences of this name, and the third column gives taxon names (if any) in the TWN list that are most similar to the one in the BEQI input file. This column may be useful to discover and correct typing errors or slightly different spelling. Please report inconvertible taxa names of Dutch benthos data to the TAXA-TWN list manager of RWS (myra.swarte@rws.nl).



## Species sensitivity values

```{r child="default_ambi.Rmd", eval=isTRUE(is.null(settings[["AMBI-file"]]))}
```

```{r child="user_ambi.Rmd", eval=!isTRUE(is.null(settings[["AMBI-file"]]))}
```


```{r}
# read AMBI's
d_sens <- readAMBI()
names(d_sens) <- c("TAXON", "AMBI")
d_sens$TAXON <- rename(x = d_sens$TAXON, from = d_twn$taxonname, to = d_twn$taxon)
d_sens <- unique(na.omit(d_sens))
if (!is.null(settings[["AMBI-file"]])) {
    d <- readAMBI(filename = settings[["AMBI-file"]])
    names(d) <- c("TAXON", "AMBI_user")
    d$TAXON <- rename(x = d$TAXON, from = d_twn$taxonname, to = d_twn$taxon)
    d <- unique(na.omit(d))
    d <- merge(x = d_sens, y = d, all = TRUE) 
    sel <- is.na(d$AMBI_user)
    d$AMBI_user[sel] <- d$AMBI[sel]
    d_sens <- data.frame(TAXON = d$TAXON, AMBI = d$AMBI_user,
                         stringsAsFactors = FALSE)
}
```


```{r results='asis'}
isMissing <- is.na(match(x = tolower(d_beqi$TAXON), table = tolower(d_sens$TAXON)))
d <- ddply(
    .data = d_beqi[isMissing, ], 
    .variables = "TAXON", 
    .fun = function(x) {
        data.frame(n = nrow(x))
    }
)
if (nrow(d) > 0L) {
    cat("\nSpecies sensitivity values for the following taxa are missing:\n",
        "(where n is the number of times each taxon is missing)\n")
    print(xtable(d), type = "html")
}
```





## Inventory of water bodies-ecotopes present in the dataset

The following ecotopes have been selected:
```{r results='asis'}
d <- table(d_beqi[, c("OBJECTID", "ECOTOPE")])
print(xtable(as.data.frame(d)), type = "html")
```




## Inventory of sample areas per water body-ecotope-year-period


### Available total sample area

```{r results='asis'}
d <- unique(
    d_beqi[, c("OBJECTID", "ECOTOPE", "YEAR", "SAMPLEID", "AREA"), drop = FALSE]
)
d <- subset(x = d, select = -SAMPLEID)
d <- dcast(
    data = d, 
    formula= ... ~ AREA, 
    fun.aggregate = sum, 
    value.var = "AREA"
)
d$total <- rowSums(d[, -(1:3), drop = FALSE])
print(xtable(d), type = "html")
```


```{r}
sel <- which(d$total  < settings$'target area'[1])
tooSmall <- NULL
if (length(sel) > 0L) {
    tooSmall <- sQuote(
        apply(
            X = d[sel, c("YEAR", "OBJECTID", "ECOTOPE"), drop = FALSE],
            MARGIN = 1,
            paste,
            collapse = "-"
        )
    )
}
    
```

The total sample area is smaller than the minimum target pool area (`r settings$'target area'[1]`) for `r length(tooSmall)` combinations of water body- ecotope-year-period.

```{r eval=length(tooSmall)>0L}
tooSmall
```




### Number of samples per sample area
```{r results='asis'}
d <- unique(
    d_beqi[, c("OBJECTID", "ECOTOPE", "YEAR", "SAMPLEID", "AREA"), drop = FALSE]
)
d <- subset(x = d, select = -SAMPLEID)
d <- dcast(
    data = d, 
    formula= ... ~ AREA, 
    fun.aggregate = length, 
    value.var = "AREA"
)
d$total <- as.integer(rowSums(d[, -(1:3), drop = FALSE]))
print(xtable(d), type = "html")

```


```{r child="genus_to_species.Rmd", eval=isTRUE(settings[["Genus to species conversion"]])}
```


```{r child="pooling.Rmd", eval=isTRUE(settings[["pool"]])}
```

```{r child="nopooling.Rmd", eval=!isTRUE(settings[["pool"]])}
```




## Indicators



```{r}
groupingVars <- c("OBJECTID", "ECOTOPE", "YEAR", "POOLRUN", "POOLID")
```

### Total abundance

The total abundance of individuals ($N$) in the data pool.


### Species richness

Species richness ($S$) is defined as the number of taxa (lowest identification level possible) per sampling unit (data pool or box core sample).


### Shannon Index

Shannon Index is given by (Shannon, 1948, p.393)
\[
    H'(X) = -\sum_i \mathrm{p}(x_i) \log_2 \left[\mathrm{p}(x_i) \right]
\]
where in our case $\mathrm{p}(x_i) = \frac{n_i}{N}$ the probability of occurrence of species $x_i$, $n_i$ is the abundance of species $x_i$ and $N$ is the total abundance, _i.e._, $N = \sum_i n_i$.

```{r}
indicators <- function(x) {

    # total abundance
    N <- sum(abundance(taxon = x$TAXON, count = x$VALUE))
    
    # species richness 
    S <- speciesRichness(taxon = x$TAXON, count = x$VALUE)
    
    # Shannon's entropy
    H <- entropy(taxon = x$TAXON, count = x$VALUE)

    # number of sampling units in the pool
    nSamplesInPool <- length(unique(x$ID))
    
    # total area of the pool
    poolArea <- sum(unique(x[, c("ID", "AREA")])$AREA)
    
    # return results
    data.frame(nSamplesInPool, poolArea, N, S, H)
}

# compute indicators
d_ind <- ddply(.data = d_beqi, .variables = groupingVars, .fun = indicators)

```



### Marine Biotic Coefficients

The table below gives an overview of the total number of taxa within each sensitivity class in each species sensitivity file for the entire data set. 

```{r results='asis'}
print(xtable(t(as.matrix(table(d_sens$AMBI)))), type = "html", 
      include.rownames = FALSE)
```

```{r}
# merge sensitivity classes with BEQI-input data
d_beqi$AMBI <- d_sens$AMBI[match(x = d_beqi$TAXON, table = d_sens$TAXON)]

# estimate frequency distribution of sensitivity classes
d <- ddply(
    .data = d_beqi, 
    .variables = groupingVars,
    .fun = function(x) {
        x$AMBI[x$VALUE == 0L] <- "azoic"
        ddply(
            .data = x, 
            .variables = "AMBI", 
            .fun = function(x) {
                data.frame(n = sum(x$VALUE, na.rm = TRUE))
            }
        )
    }
)
```


The average percentage of the total abundance without an AMBI classification is given below:

```{r results='asis'}
tmp <- ddply(
    .data = d, 
    .variables = c("OBJECTID", "ECOTOPE", "YEAR"),
    .fun = function(x) {
        index <- which(is.na(x$AMBI))
        if (length(index) == 0L) {
            p <- 0
        } else {
            p <- 100 * sum(x$n[index])/sum(x$n)
        }
        data.frame(missing = p, warning = ifelse(p > 20, "!", ""))
    }
)
print(xtable(tmp), type = "html")
```

Note that `r sum(tmp$warning != "")` percentages exceed 20.

Borja et al. (2000) introduced the Biotic Coefficient, defined as:
\[
c_b = \mathbf{w}^\prime \mathbf{f} = \sum_{i=1}^5 w_i f_i
\]
where $\mathbf{w}=[0, 1.5, 3, 4.5, 6]^\prime$ is a vector of weights and $\mathrm{f}$ is a vector of length 5 containing the fractions of species in the sensitivity classes (I, II, III, IV, V). The prime $\prime$ denotes vector transpose.

In order to estimate the biotic coefficients, `r sum(is.na(d$AMBI))` taxa with unknown sensitivity classes will be excluded from analysis.


```{r}
# compute biotic coefficients
d <- d[!is.na(d$AMBI), ]
d <- ddply(
    .data = d, 
    .variables = groupingVars,
    .fun = function(x) {
        if (all(x$n == 0L)) {
            # handle samples that are fully azoic
            biotic_coefficient <- NA_real_
        } else {
            # handle samples that contain azoic subsamples, 
            # but are not azoic as a whole. Azoic subsamples 
            # have to be removed prior to further analysis.
            x <- x[x$n != 0L, , drop = FALSE]

            # compute biotic coefficients
            f <- x$n/sum(x$n)
            w <- c(I = 0, II = 1.5, III = 3, IV = 4.5, V = 6)
            w <- w[as.character(x$AMBI)] # alignment
            biotic_coefficient <- sum(w * f)
        }
        data.frame(AMBI = biotic_coefficient)
    }
)

d_ind <- merge(x = d_ind, y = d, all = TRUE)

```


### Infaunal Trophic Index

The infaunal trophic index (ITI) is calculated as:
\[
    \mathrm{ITI} = \frac{1}{3}\sum_{i=1}^3 (4-i) f_i
\]
where $f_i$ is the fraction of species in class $i$, where 
- class 1 are suspension feeders (highest quality);
- class 2 are interface feeders
- class 3 are surface deposit feeders and
- class 4 are subsurface deposit feeders (lowest quality). 

See Gittenberger & van Loon (2013) for more information.


```{r eval=!is.null(settings[["ITI-file"]])}
# read ITI data
d <- readITI(settings[["ITI-file"]])
names(d) <- c("TAXON", "ITI")

# convert TAXON names to preferred names
d$TAXON <- rename(x = d$TAXON, from = d_twn$taxonname, to = d_twn$taxon)
d <- na.omit(d)

# add ITI score to BEQI2-data
index <- match(x = d_beqi$TAXON, table = d$TAXON)
d_beqi$ITI <- d$ITI[index]

d <- ddply(
    .data = d_beqi, 
    .variables = groupingVars, 
    .fun = function(x) {
        # estimate frequency
        f <- tapply(X = x$VALUE, INDEX = x$ITI, FUN = sum)
        f <- f / sum(f)
        
        # estimate and return ITI
        w <- (4-(1:4)) / 3
        names(w) <- c("I", "II", "III", "IV")
        w <- w[names(f)] # alignment
        data.frame(ITI = sum(w * f))
    
})


d_ind <- merge(x = d_ind, y = d, all = TRUE)

```



```{r eval=!is.null(settings[["FIBI-file"]])}

### FIsheries Biotic Index 
#The FIsheries Biotic Index (FIBI) is calculated as:
#\[
#    \mathrm{FIBI} = \frac{1}{3}\sum_{i=1}^3 (4-i) f_i
#\]
#where $f_i$ is the fraction of species in class $i$, where
#- class 1 is very sensitive to fisheries (highest quality);
#- class 2 is sensitive to fisheries
#- class 3 is insensitive to fisheries and
#- class 4 are second-order opportunistic species (lowest quality). 


# read FIBI data
d <- readFIBI(settings[["FIBI-file"]])
names(d) <- c("TAXON", "FIBI")

# convert TAXON names to preferred names
d$TAXON <- rename(x = d$TAXON, from = d_twn$taxonname, to = d_twn$taxon)
d <- na.omit(d)

# add FIBI score to BEQI2-data
index <- match(x = d_beqi$TAXON, table = d$TAXON)
d_beqi$FIBI <- d$FIBI[index]

d <- ddply(
    .data = d_beqi, 
    .variables = groupingVars, 
    .fun = function(x) {
        # estimate frequency
        f <- tapply(X = x$VALUE, INDEX = x$FIBI, FUN = sum)
        f <- f / sum(f)
        
        # estimate and return FIBI
        w <- (4-(1:4)) / 3
        names(w) <- c("I", "II", "III", "IV")
        w <- w[names(f)] # alignment
        data.frame(FIBI = sum(w * f))
})


d_ind <- merge(x = d_ind, y = d, all = TRUE)

```














### Percentiles


Percentiles for each indicator are given below. In addition, the number of samples (n) used to estimate these percentiles is provided. The percentiles have been calculated for the period `r paste(range(d_beqi$YEAR), collapse = "-")` and months `r paste(range(settings[["months"]]), collapse = "-")`
```{r results='asis'}
d <- ddply(
    .data = d_ind,
    .variables = c("OBJECTID", "ECOTOPE"),
    .fun = function(x) {
        x <- x[, !(names(x) %in% c("OBJECTID", "ECOTOPE", "YEAR", "POOLRUN", 
                                   "poolArea", "POOLID", "nSamplesInPool", "N"))]
        n <- sapply(X = x, FUN = function(x){sum(!is.na(x))})
        x <- ldply(.data = x, .fun =  quantile, probs = c(0, 1, 95, 99, 100) / 100, na.rm = TRUE)
        names(x)[names(x) == ".id"] <- "indicator"
        x$n <- n
        x
    }
)
print(xtable(d), type = "html")

```



### Ecological Quality Ratios

The following Ecological Quality Ratios (EQRs) will be calculated

- For species richness:
\[
    EQR(S) = \frac{S_\mathrm{ass} - S_\mathrm{bad}}{S_\mathrm{ref} - S_\mathrm{bad}}
\]
- For Shannon's Index:
\[
    EQR(H) = \frac{H_\mathrm{ass} -  H_\mathrm{bad}}{H_\mathrm{ref} -  H_\mathrm{bad}}
\]
- For AMBI
\[
    EQR(AMBI) = \frac{\mathrm{AMBI}_\mathrm{bad} -  \mathrm{AMBI}_\mathrm{ass}}{\mathrm{AMBI}_\mathrm{bad} - \mathrm{AMBI}_\mathrm{ref}} = \frac{6 - \mathrm{AMBI}_\mathrm{ass}}{6}
\]
- For BEQI-2
\[
    EQR(BEQI2) = \frac{EQR(S) + EQR(H) + EQR(AMBI)}{3} + c
\]
where $c$ is an additive correction term, which is given as `CORR` in the table below.
<br>

The 'bad' and 'reference' values can be found in the table below. This table is a copy of
`r sQuote(settings[["ER-file"]])` as specified in the settings file.
```{r results='asis'}
d_erf <- readERF(filename = settings[["ER-file"]])
print(xtable(d_erf), type = "html", include.rownames = FALSE)
```


```{r}
d_erf$OBJECTID <- harmonize(d_erf$OBJECTID)
d_erf$ECOTOPE  <- harmonize(d_erf$ECOTOPE)
sel <- !(names(d_erf) %in% c("OBJECTID", "ECOTOPE", "RELAREA"))
names(d_erf)[sel] <- paste("erf", names(d_erf)[sel], sep = "_")
d_ind <- merge(x = d_ind, y = d_erf, all.x = TRUE, all.y = FALSE)
```


```{r eval=any(is.na(d_ind$erf_SREF)), results='asis'}
cat("WARNING: No reference data available for:\n\n")
print(
    xtable(d_ind[is.na(d_ind$erf_SREF), c("OBJECTID", "ECOTOPE"), drop = FALSE]), 
    type = "html"
)

d_ind <- d_ind[!is.na(d_ind$erf_SREF), ]
```

```{r}
d_ind$S_EQR <- eqr(
    x = d_ind$S, 
    bad = d_ind$erf_SBAD, 
    good = d_ind$erf_SREF
)

d_ind$H_EQR <- eqr(
    x = d_ind$H, 
    bad = d_ind$erf_HBAD, 
    good = d_ind$erf_HREF
)

# estimate EQR of ITI (optional)
if (!is.null(settings[["ITI-file"]])) {
    d_ind$ITI_EQR <- eqr(
        x = d_ind$ITI, 
        bad = d_ind$erf_ITIBAD, 
        good = d_ind$erf_ITIREF
    )
}

# estimate EQR of FIBI (optional)
if (!is.null(settings[["FIBI-file"]])) {
    d_ind$FIBI_EQR <- eqr(
        x = d_ind$FIBI, 
        bad = d_ind$erf_FIBIBAD, 
        good = d_ind$erf_FIBIREF
    )
}

# estimate EQR of AMBI
d_ind$AMBI_EQR <- eqr(
    x = d_ind$AMBI, 
    bad = d_ind$erf_AMBIBAD,
    good = d_ind$erf_AMBIREF
)


# estimate BEQI2-EQR
d_ind$BEQI2_EQR <- rowMeans(d_ind[, c("S_EQR", "H_EQR", "AMBI_EQR")])

# apply inter-calibration correction of BEQI2
d_ind$BEQI2_EQR <- d_ind$BEQI2_EQR + d_ind$erf_CORR

```

---

## Results

The results are averaged to water body-ecotope-year-period combinations.

The table below lists all results aggregated by OBJECTID, ECOTOPE and YEAR.

```{r results='asis',  eval=nrow(d_ind)>0L}
# aggregate by pool run
v <- c("OBJECTID", "ECOTOPE", "YEAR")
w <- c(v, "nSamplesInPool", "poolArea", "RELAREA",
       "N", "S", "H", "AMBI", 
       if(is.null(settings[["ITI-file"]])){NULL} else{"ITI"},
       if(is.null(settings[["FIBI-file"]])){NULL} else{"FIBI"},
       "S_EQR", "H_EQR", "AMBI_EQR",
       if(is.null(settings[["ITI-file"]])){NULL} else{"ITI_EQR"},
       if(is.null(settings[["FIBI-file"]])){NULL} else{"FIBI_EQR"},
       "BEQI2_EQR"
)

# aggregate over pools runs and pooled samples
d <- ddply(
    .data = d_ind[, w], 
    .variables = v,
    .fun = function(x) {
        colMeans(x[, !(names(x) %in% v)])
    }
)

print(xtable(d), type = "html")
write.csv(
    x = d, 
    file = file.path(
        dirname(settings[["output-file"]]), 
        paste0("ECOTOPE_", basename(settings[["output-file"]]))
    ),
    row.names = FALSE
)

```


<br>
The table below lists all results aggregated (ecotope area-weighted) by OBJECTID and YEAR.




```{r results='asis'}
# aggregate by pool run
v <- c("OBJECTID", "YEAR")
w <- c(v, "RELAREA", "BEQI2_EQR")
d <- ddply(
    .data = d[, w], 
    .variables = v,
    .fun = function(x) {
        x <- x[, !(names(x) %in% v)]
        # weighted averaging (safe version, also works when RELAREA doesn't sum to 1)
        x <- colSums(x * x$RELAREA) / sum(x$RELAREA)
        x[names(x) != "RELAREA"]
    }
)
print(xtable(d), type = "html")
write.csv(
    x = d, 
    file = file.path(
        dirname(settings[["output-file"]]), 
        paste0("OBJECTID_", basename(settings[["output-file"]]))
    ),
    row.names = FALSE
)

```



---

    
    

## References

Borja, A., J. Franco and V. P&eacute;rez, 2000. A Marine Biotic Index to Establish the
Ecological Quality of Soft-Bottom Benthos Within European Estuarine and Coastal Environments. Marine Pollution Bulletin 40:1100-1114

Gittenberger A. and  W. van Loon, 2013. Sensitivities of marine macrozoobenthos to environmental pressures in the Netherlands. Nederlandse Faunistische Mededelingen, 41 (2013) 79-112.

Shannon, C. E., 1948. A Mathematical Theory of Communication. Bell System Technical Journal 27: 379–423.

van Loon, W.M.G.M. and A.J. Verschoor, 2012. Benthic Ecosystem Quality Index 2: Application to Dutch Marine Benthos Data from the Period 1990 - 2010, Report RWS Waterdienst.

van Loon, W.M.G.M., A.J. Verschoor and A. Gittenberger, 2012. Benthic Ecosystem Quality Index 2: Design and Calibration of the BEQI-2 WFD Metric for Marine Benthos in Transitional Waters, Report RWS Waterdienst.


---
## Acknowledgements
 
Dr. Angel Borja and the AZTI-TECHNALIA, Spain, are acknowledged for their kind permission to use the AMBI species list by the BEQI-2 tool only.
 
If you want to use the AMBI tool and species list stand alone, you can download these from the website ambi.azti.es.
 
For more information on the AMBI software and species list, please contact Dr. Angel Borja, email: aborja@azti.es.



---


## Session information 

```{r}
sessionInfo()
```
